\mythesistitleNOTSYNPOSIS

### 2012-01-05 src/020chap2.tex
Chapter 2 edited. See diff for details. Too many to remember. or maybe too little.

@@

\begin{definition}{\emph{Matrices that define
      graphs.\cite[Def.~2.4]{d08phd}}} %
  \label{def:graphmatrices} %
  Let $G$ and $H$ be defined as follows. $G = (V,E_G)$ is a graph with
  vertex set $V = \{v_i \mid i \in [n]\}$ and edge set $E_G \subseteq
  \{(v_i,v_j) \mid i, j \in [n]\}$ such that $|E_G| = m$. $H = (A, B,
  E_H)$ is a bipartite graph with partitions $A = \{a_i \mid i \in
  [n_a]\}$ and $B = \{b_i \mid i \in [n_b]\}$.
  \begin{enumerate}[{\ref{def:graphmatrices}}--i.]
  \item \emph{Adjacency matrix} of $G$ is the symmetric $n \times n$
    binary matrix $M$ with $m_{i,j} = \un$ \iff $(v_i,v_j) \in E_G$
    for all $i,j \in [n]$.
  \item \emph{Augmented adjacency matrix} of $G$ is obtained from its
    adjacency matrix by setting all main diagonal elements to \un, \ie
    $m_{i,i} = \un$ for all $i \in [n]$.
  \item \emph{Maximal clique matrix} or \emph{vertex-clique incidence
      matrix} of $G$ is the $n \times k$ binary matrix $M$ with
    $m_{i,j} = \un$ \iff $v_i \in C_j$ for all $i \in [n], j \in [k]$
    where $\{C_j \mid j \in [k]\}$ is the set of maximal cliques of
    $G$.
  \item \emph{Half adjacency matrix} of $H$ is the $n_a \times n_b$
    binary matrix $M$ with $m_{i,j} = \un$ \iff $(a_i, b_j) \in E_H$
  \end{enumerate}
\end{definition}

@@

%\hrule
\begin{definition}{\emph{Graphs that relate to
      COP.\cite[Def.~2.5]{d08phd}}} %
   \label{def:graphwithcop} %
   Let $G$ be a graph and $H$ be a bipartite graph.

   \begin{enumerate}[\ref{def:graphwithcop}--i.]
   \item $G$ is \emph{convex-round} if its adjacency matrix has the
     \CROP.
   \item $G$ is \emph{concave-round} if its augmented adjacency matrix
     has \CROP. \tnote{cite BHY00} \tnote[minor]{add CROP to glossary}
   \item $G$ is an \emph{interval graph} if its vertices can be mapped
     to intervals on the real line \stt two vertices are adjacent \iff
     their corresponding intervals overlap \tnote{cite Ben59, Haj57}.
     $G$ is an interval graph \iff its maximal clique matrix has COP
     due to the result by \cite{gh64} which states that the maximal
     cliques of interval graph $G$ can be linearly ordered \stt for
     all $v \in V(G)$, cliques containing $v$ are consecutive in the
     ordering.
   \item $G$ is a \emph{unit interval graph} if it is an interval
     graph \stt all intervals have the same length.
   \item $G$ is a \emph{proper interval graph} if it is an interval
     graph \stt no interval properly contains another.
   \item $G$ is a \emph{circular-arc graph} if its vertices can be
     mapped to a set of arcs on a circle \stt two vertices are
     adjacent \iff their corresponding arcs overlap.
   \item $H$ is \emph{convex bipartite} if its half adjacency matrix
     has \COP on either rows or columns.
   \item $H$ is \emph{biconvex bipartite} or \emph{doubly
       convex}\cite{yc95} if its half adjacency matrix has \COP on
     both rows and columns.
   \item $H$ is \emph{circular convex} if its half adjacency matrix
     has \CROP.
   \end{enumerate}
\end{definition}
%\hrule

%%%%%%                                              %%%%%%
% [pressing]{\ \ \ \ \  is the second level of [pressing]%
%%%%%%                                              %%%%%%
\tnote[pressing]{ADD: As it will be described in detail later in this
  document, isomorphism of certain % classes of graphs, namely chordal
  % graphs, have a close relationship with consecutive-ones property
  % and generalizations of it.  This is perhaps because of how closely
  % COP of a matrix relates to properties of graphs derived from
  % matrices as seen inz the following results.
} \tnote[pressing]{\ \ \ \ \ ADD: peo exists iff chordal. lexicographic BFS
  [tag:chordalGraph]} %
\tnote[pressing]{\ \ \ \ \ ADD: A well known result in %perfect graph theory is that
  % the maximal cliques of an interval graph $G$ can be linearly
  % ordered such that for all $v \in V(G)$, cliques containing $v$ are
  % consecutive in the ordering cite~{gh64}. This clearly means that a
  % graph $G$ is an interval graph if and only if
}%
\tnote[pressing]{\ \ \ \ \ verify from paper the statement of claim.}%
\tnote[pressing]{maximal clique vertex incidence matrix
  of% $G$ has COP.  Also, maximal cliques of any
  % chordal graph can be enumerated in polytime $O(m+n)$
} \tnote[pressing]{\ \ \ \ \ citation?!!}%
\tnote[pressing]{cite~{fg65} uses these results to give the first polynomial
  time algorithm for COT.}%
\tnote[pressing]{\ \ \ \ \ check. how do they use it?}%
\tnote[pressing]{ A bipartite graph is
  convex% (on $R$) if and only if its half adjacency matrix has COP
  % on rows.  The results in cite~{bl76} on COT are based on the
  % result that interval graphs are AT-free chordal graphs.
}%
\tnote[pressing]{\ \ \ \ \ the latter being Tucker's?}%
\tnote[pressing]{\ \ \ \ \ (2) TBD survey
  -- % see blue notes (in notebook) under Graph Isomorphism. namely
  % citations in cite:aas93 (3) a brief on heirarchy of chordal
  % graphs, path graphs, interval graphs, peo, clique tree etc. and
  % results. (4) interval graphs are incomparability graphs. see
  % Golumbic. [tag:classification] (5) finding min length hole in
  % bipartite graph is polynomial time [Sec 3.3.2 in cite:d08phd] (6)
  % theorem 2.2 in cite:d08phd - $G$ is union of v.d. caterpillars iff
  % $M$ has COP in rows. $M$ is edge vertex incidence matrix $(*,2)$
  % of $G$
}%

\section{Matrices with COP}
\label{sec:surveycoptest}
\tnote{Expand on ref:sec:copmatrices}

Figure~\ref{fig:cop-matrix} shows examples of consecutive-ones
property. \tnote[important]{move to general def section? if so, decide
how to cross ref with repetition for completeness of chapter.}

\begin{figure}[t] %[htbp]
  \centering

  {\figtabsize
    \begin{tabular}[h]{l|lcccl}
      $M_1$: & $M_1'$: &&&& $M_2$:\\
      &&&&&\\
      \begin{tabular}[h]{llll}
        $c_1$ & $c_2$ &$c_3$ &$c_4$\\
        &&&\\
        \un & 0   & \un & 0\\
        0   & \un & 0   & \un \\
        \un & 0   & 0   & \un
      \end{tabular}
      &
      \begin{tabular}[h]{llll}
        $c_3$ &$c_1$ &$c_4$& $c_2$\\
        &&&\\
        \un & \un & 0 & 0\\
        0 & 0 & \un & \un \\
        0 & \un & \un & 0
      \end{tabular}
      &&&&
      \begin{tabular}[h]{llll}
        $d_1$ & $d_2$ &$d_3$ &$d_4$\\
        % &&&\\
        &&&\\
        \un & \un & 0 & 0\\
        0 & \un & \un & 0 \\
        0 & \un & 0 & \un 
      \end{tabular}
    \end{tabular}
  }

  \caption[\figtabsize Matrices with and without COP.]{\figtabsize
    Matrices with and without COP. $M_1$ has COP because by permuting
    its columns, $c_1$-$c_4$, one can obtain $M_1'$ where the {\un}s
    in each row are consecutive. $M_2$, however, does not have COP
    since no permutation of its columns, $d_1$-$d_4$, will arrange
    {\un}s in each row consecutively \cite{d08phd}.}

  \label{fig:cop-matrix}
\end{figure}

\section{Optimization problems in COP}
\label{sec:surveycopopt}
\tnote{Expand on ref:sec:optcop}

\section{********* COP in Relational Database Model}
\label{sec:surveyrdbm}
\tnote{Expand on sec:apprdbm} \tnote{(set systems theme)}

\section{********* COP in Graph Isomorphism}
\label{sec:surveygraphiso}
\tnote{Expand on sec:appgraphiso} \tnote{(canonization theme)}

\section{********* Certifying Algorithms}
\label{sec:surveycertalgo}
\tnote{Expand on sec:appcertalgo} \tnote{ (certification McC04 theme)}

\section*{-- REFERENCE CONTENT --}
%%%%%%%%%%%%% listing package for pretty formatting BEGIN
% \lstloadlanguages{C}
% \lstset{language=C, commentstyle=\scriptsize,
%   numberstyle=\tiny, numbers=left,               % Line numbers
%   stepnumber=2,numbersep=5pt,firstnumber=10, 
%   showstringspaces=false,                    % No explicit spaces
%   emph={printf},emphstyle=\underbar,      % Specific formatting
%   emph=[2]{sum},emphstyle=[2]\color{blue},      % Sp format (multiple)
%   emph=[3]{i},emphstyle=[3]\color{red}      % Sp format (multiple)
%   }

{%\fontfamily{cmvtt}\selectfont
\color{blue}
\subsection{Matrices with COP}
\label{sec:copmatrices}
As seen earlier, the interval assignment problem (illustrated as the
course scheduling problem in Section~\ref{sec:problem}), is a special
case of the problem we address in this thesis, namely the tree path
labeling problem (illustrated as the \illustrationproblem). The
interval assignment problem and COP problem are equivalent
problems. In this section we will see some of the results that exists
in the literature today towards solving the COP problem and
optimization problems surrounding it.

Recall that a matrix with COP is one whose rows (columns) can be
rearranged so that the {\un}s in every column (row) are in consecutive
rows (columns). COP in binary matrices has several practical applications
in diverse fields including scheduling \cite{hl06}, information
retrieval \cite{k77} and computational biology \cite{abh98}.  Further,
it is a tool in graph theory \cite{mcg04} for interval graph
recognition, characterization of Hamiltonian graphs, planarity testing
\cite{bl76} and in integer linear programming \cite{ht02,hl06}.

The obvious first questions after being introduced to the consecutive
ones property of binary matrices are if COP can be detected
efficiently in a binary matrix and if so, can the COP permutation of
the matrix also be computed efficiently?  Recognition of COP in a
binary matrix is polynomial time solvable and the first such algorithm
was given by \cite{fg65}.  A landmark result came a few years later
when \cite{at72} discovered the families of forbidden submatrices that
prevent a matrix from having COP and most, if not all, results that
came later were based on this discovery which connected COP in binary
matrices to convex bipartite graphs. In fact, the forbidden
submatrices came as a corollary to the discovery that convex bipartite
graphs are AT-free in \cite{at72}\tnote{check}. The first linear
time algorithm for COP testing (COT) was invented by \cite{bl76} using
a data structure called PQ trees.  Since then several COT algorithms
have been invented -- some of which involved variations of PQ trees
\cite{mm96,wlh01,mcc04}, some involved set theory and ICPIA
\cite{wlh02,nsnrs09}, parallel COT algorithms\cite{as95,bs03,ly91} and
certifying algorithms\cite{mcc04}.

The construction of PQ trees in \cite{bl76} draws on the close
relationship of matrices with COP to interval graphs. A PQ tree of a
matrix is one that stores all row (column) permutations of the matrix
that give the COP orders (there could be multiple orders of rows or
columns) of the matrix. This is constructed using an elaborate linear
time procedure and is also a test for planarity\tnote{check check
  check. both interval graph and planarity in this paper?}.  PQR trees
is a generalized data structure based on PQ trees \cite{mm96,mpt98}.
\cite{tm05} describes an improved algorithm to build PQR
trees. \tnote{improv in terms of what?}\cite{wlh02} describes the
simpler algorithm for COT. Hsu also invented PC trees
\cite{wlh01}\tnote{This result first appeared inproc ISAAC92}
which is claimed to be much easier to implement. \cite{nsnrs09}
describes a characterization of consecutive-ones property solely based
on the cardinality properties of the set representations of the
columns (rows); every column (row) is equivalent to a set that has the
row (column) indices of the rows (columns) that have one entries in
this column (row). This is interesting and relevant, especially to
this thesis because it simplifies COT to a great degree. \tnote{it
  reduces the solution search space. fill in the blanks.}

\cite{mcc04} describes a different approach to COT. While all previous
COT algorithms gave the COP order if the matrix has the property but
exited stating negative if otherwise, this algorithm gives an evidence
by way of a certificate of matrix even when it has no COP. This
enables a user to verify the algorithm's result even when the answer
is negative. This is significant from an implementation perspective
because automated program verification is hard and manual verification
is more viable. Hence having a certificate reinforces an
implementation's credibility. Note that when the matrix {\em has} COP,
the COP order is the certificate.  The internal machinery of this
algorithm is related to the weighted betweenness problem
addressed\tnote{in what way??} in \cite{co98}.  \tnote{expand
  on the COP order graph creation and it having to be bipartite for M
  to have COP. and thus an odd cycle being an evidence of no COP.}
\tnote{ where should this go?: (1) cite|jlm97 (application of PQ
  trees in graphics). (2) helly's theorem citation
  19XXdgk-Hellystheorem-Danzer-Gruenbaum-Klee}

\subsection{Optimization problems in COP}
\label{sec:optcop}

So far we have been concerned about matrices that have the consecutive
ones property. However in real life applications, it is rare that data
sets represented by binary matrices have COP, primarily due to the
noisy nature of data available. At the same time, COP is not arbitrary
and is a desirable property in practical data representation
\cite{co98,jkckv04,k77}. In this context, there are several
interesting problems when a matrix does not have COP but is ``close''
to having COP or is allowed to be altered to have COP. These are the
optimization problems related to a matrix which does not have
COP. Some of the significant problems are surveyed in this section.

@@

\cite{jkckv04} discusses the use of matrices with almost-COP (instead
of one block of consecutive {\un}s, they have $x$ blocks, or {\em
  runs}, of consecutive {\un}s and $x$ is not too large) in the
storage of very large databases.  The problem is that of reordering of
a binary matrix such that the resulting matrix has at most $k$ runs of
{\un}s. This is proved to be NP hard using a reduction from the
Hamiltonian path problem.\tnote{Theorem 2.1 in jkckv} \tnote{(1) A
  connection of COP problem to the travelling salesman problem is also
  introduced. what does this mean? -- COP can be used as a tool to
  reorder $0.5T \le runs(M) \le T$. (2) The optimization version of
  the $k$-run problem, i.e. minimization of number of blocks of ones
  is proven to be NP complete by cite:k77}\tnote{are these two the
  same?} \tnote{what is the reduction?} \tnote{other problems similar
  to COP -- cite:ckl96 (ILP, circ ones, one drop) -- cite:th98
  (generalization of COP - minimax, biotonic column) Tucker} }

@@

>>>>

\definecolor{gray}{rgb}{0.8,0.8,0.8}
\def\tblhcolor {gray}

@@

% \begin{definition}{\emph{Matrices that define
%       graphs.\cite[Def.~2.4]{d08phd}}} %
\begin{definition}
  {\em Matrices that define graphs. \cite[Def.~2.4]{d08phd}.}  
  \label{def:graphmatrices} 
  Let $G$ and $H$ be defined as follows. $G = (V,E_G)$ is a graph with
  vertex set $V = \{v_i \mid i \in [n]\}$ and edge set $E_G \subseteq
  \{(v_i,v_j) \mid i, j \in [n]\}$ such that $|E_G| = m$. $H = (A, B,
  E_H)$ is a bipartite graph with partitions $A = \{a_i \mid i \in
  [n_a]\}$ and $B = \{b_i \mid i \in [n_b]\}$.
  \begin{enumerate}[{\ref{def:graphmatrices}}--i.]
  \item \emph{Adjacency matrix} of $G$ is the symmetric $n \times n$
    binary matrix $M$ with $m_{i,j} = \un$ \iff $(v_i,v_j) \in E_G$
    for all $i,j \in [n]$.
  \item \emph{Augmented adjacency matrix} of $G$ is obtained from its
    adjacency matrix by setting all main diagonal elements to \un, \ie
    $m_{i,i} = \un$ for all $i \in [n]$.
  \item \emph{Maximal clique matrix} or \emph{vertex-clique incidence
      matrix} of $G$ is the $n \times k$ binary matrix $M$ with
    $m_{i,j} = \un$ \iff $v_i \in C_j$ for all $i \in [n], j \in [k]$
    where $\{C_j \mid j \in [k]\}$ is the set of maximal cliques of
    $G$.
  \item \emph{Half adjacency matrix} of $H$ is the $n_a \times n_b$
    binary matrix $M$ with $m_{i,j} = \un$ \iff $(a_i, b_j) \in E_H$.
  \end{enumerate}
\end{definition}

\figgraphmatrices

@@

\begin{definition}{\emph{Graphs that relate to
      COP.\cite[Def.~2.5]{d08phd}}} %
  \label{def:graphwithcop} %
  Let $G$ be a graph and $H$ be a bipartite graph.

  \begin{enumerate}[\ref{def:graphwithcop}--i.]
  \item $G$ is \emph{convex-round} if its adjacency matrix has the
    \CROP.
  \item \label{def::concave-round} $G$ is \emph{concave-round} if its
    augmented adjacency matrix has \CROP. \tnote{cite BHY00}
    \tnote[minor]{add CROP to glossary}
  \item $G$ is an \emph{interval graph} if its vertices can be mapped
    to intervals on the real line \stt two vertices are adjacent \iff
    their corresponding intervals overlap \tnote{cite Ben59, Haj57}.
    $G$ is an interval graph \iff its maximal clique matrix has COP
    \cite{fg65}\endnote{This follows \cite{gh64} which states that the
      maximal cliques of interval graph $G$ can be linearly ordered
      \stt for all $v \in V(G)$, cliques containing $v$ are
      consecutive in the ordering \cite[Th. 8.1]{mcg04}.}
    \begin{enumerate}[a.]
    \item $G$ is a \emph{unit interval graph} if it is an interval
      graph \stt all intervals have the same length.\endnotemark[2]
    \item $G$ is a \emph{proper interval graph} if it is an interval
      graph \stt no interval properly contains another.\endnote{The
        set of unit interval graphs and the set of proper interval
        graphs are the same} \tnote{cite {rob69,gar07} in endnote. pg
        33 dom}
    \end{enumerate}
 \item $G$ is a \emph{circular-arc graph} if its vertices can be
    mapped to a set of arcs on a circle \stt two vertices are adjacent
    \iff their corresponding arcs overlap.\tnote[pressing]{how is
      CO/ROP related?}
  \item \label{def::convexbi} $H$ is \emph{convex bipartite on columns (rows)} if its half adjacency matrix
    has \COP on rows (columns).
  \item $H$ is \emph{biconvex bipartite} or \emph{doubly
      convex}\cite{yc95} if its half adjacency matrix has \COP on both
    rows and columns.
  \item $H$ is \emph{circular convex} if its half adjacency matrix has
    \CROP.
  \end{enumerate}
\end{definition}

Interval graphs and circular-arc graphs have a long history in
research. The interest around them is due to their very desirable
property that several problems that are NP-hard\tnote[minor]{all NPC
  problems are NPH yes?} on general graphs are not so in these graph
classes, \ie they are polynomial time solvable on them. In a similar
fashion, a lot of problems that are hard on general matrices have
efficient solutions on matrics with \COP or \CROP \cite[more citations
pg.\,33]{d08phd}.

Table~\ref{tab:graphmatrices} summarises the way these graphs are
characterized by their matrices having \COP or \CROP.
Our focus in this chapter (and thesis) is mainly \COP and 
having seen how useful \COP is in identifying or characterizing many
types of graphs, we will now see results that study recognition of \COP
in matrices in the following section.

\temptext{ tab~\ref{tab:graphmatrices} -  have an abridged version
  of table 2.1 in dom. see notes for the possibility of a
  ``circle diagram''.} 
\begin{table}[htbp]
  \centering
 \caption{\figtabsize Graph matrices \temptext{PLACEHOLDER}}
  \label{tab:graphmatrices}
\end{table}

\clearpage      % Print pending floats

\section{Matrices with COP}
\label{sec:surveycoptest}

The most important questions with respect to a particular 
property desired in a structure/object are perhaps the following.
\begin{itemize}%[]
\singlespacing
\item Does the desired property exist in the given input?
\item If the test is affirmative, what is a certificate of the affirmative?
\item If the test is negative, what are the optimization possibilities
  for the property in the input? In other words, how close to having the
  property can the input be?
\item If the test is negative, what is a certificate of the negative?
\end{itemize}

In this section and the rest of the chapter we see results that shaped the
correponding areas respectively for \cop in binary matrices.

\begin{enumerate}[a.]
\singlespacing
\item \label{q:testcop} Does a given binary matrix have \COP?
\item \label{q:copperm} What is the \COP permutation for the given matrix with \COP?
\item \label{q:copopt} What are the optimizations possible and practically useful on
  the given matrix without \COP?
\item \label{q:nocopcert} If algorithm for (\ref{q:testcop}) returns \textbf{false}, can a
  certificate for this be computed?
\end{enumerate}

Without doubt, besides computing answers to these questions, we are
interested in the efficiency of these computations in terms of
computational complexity theory. Results towards
questions~(\ref{q:testcop}) and (\ref{q:copperm}) are surveyed in this
section. Those for question~(\ref{q:copopt}) are discussed in
Section~\ref{sec:surveycopopt} and question~(\ref{q:nocopcert}) is discussed
in Section~\ref{sec:surveycertalgo}.

It may be noted that one way to design an algorithm to test for \COP
is by deriving one from any interval graph recognition algorithm using the
result {\sc hmpv00}\tnote{cite hmpv00} \tnote{in endnote put the theorem 2.7 - dom
  pg 43} \cite{d08phd} which demonstrates how such a derivation can be
done. However, this does not necessarily yield an efficient
algorithm. We will see results that directly solve the problem on
matrices since it is known that questions~(\ref{q:testcop}) and
(\ref{q:copperm}) stated above for \COP are efficiently solvable.
Table~\ref{tab:cophistory} gives a snapshot of these results.

\begin{table}[htbp]
  \centering
   \includegraphics[scale=0.5]{../img/cophistory.png} % TEMPORARY

  \caption[\figtabsize A brief history of \COP research]{\figtabsize A
    brief history of \COP research \temptext{ OLD IMG - PLACEHOLDER - A LATEX
    TABLE WILL REPLACE THIS + WILL BE UPDATED }}
  \label{tab:cophistory}
\end{table}

\subsection{Tucker's forbidden submatrices for \COP}

% \change{While it is remarkable that the first polynomial time
%   algorithm for \COP testing came in }{The first polynomial time
%   algorithm for \COP testing came in }\cite{fg65} \change{,}{.}
% \change{it is safe to say that the most}{A few years later, a very}
% significant result in understanding \COP came \remove{a few years
%   later} from Tucker which gave a combinatorial (negative)
% characterization of matrices with COP \cite{at72}.
A polynomial time algorithm for \COP testing was seen for the first
time in \cite{fg65} which uses overlapping properties of columns with
\un s. A few years later, a deeply significant result based on very
different ideas in understanding \COP came from Tucker which gave a
combinatorial (negative) characterization of matrices with COP
\cite{at72}. This result influenced most of the \COP results that
followed in the literature \annote[pressing]{including linear time
  algorithms}{did it? which ones?} for \COP recognition.

\cite{at72} discovered certain forbidden structures for convex
bipartite graphs\endnote{The terminology in \cite{at72} differs. It
  uses the term {\em graphs with $V_1$-consecutive arragement} instead
  of {\em convex bipartite graphs}.} and by definition of this graph
class, this translates to a set of forbidden submatrices for matrices
with \cop.  The following are the theorems from \cite{at72} that
acheived this characterization.

Theorem~\ref{th:tuckeratfree} proves that convex bipartite graphs
cannot have {\em asteroidal triples}\endnote{If $G = (V,E)$ is a
  graph, a set of three vertices from $V$ form an {\em asteroidal
    triple} if between any two of them there exists a path in $G$ that
  does not contain any vertex from the closed neighborhood of the
  third vertex.} contained in the corresponding vertex
partition\endnote{The partition corresponds to columns (rows) if its
  half adjacency matrix has \COP columns (rows).}.
Theorem~\ref{th:tuckerforbidden} shows what are the structures in a
bipartite graph that force one of its vertex partitions to have
asteriodal triples -- in other words, it identifies the subgraphs that
prevent the graph from being convex bipartite.

\begin{theorem}
  (\cite[Th.~6]{at72}, \cite[Th.~2.3]{d08phd})\\
  A bipartite graph $G = (V_1, V_2, E)$ is convex bipartite on
  columns\endnote{Abridged to match terminology adopted in this
    document. See previous note.} \iff $V_1$ contains no asteroidal
  triple of $G$.
  \label{th:tuckeratfree}
\end{theorem}

\begin{theorem}
  (\cite[Th.~7]{at72}, \cite[Th.~2.4]{d08phd})\\
  In a bipartite graph $G = (V_1, V_2, E)$ the vertex set $V_1$
  contains no asteroidal triple \iff $G$ contains none of the graphs
  $G_{I_k}$, $G_{II_k}$, $G_{III_k}$ (with $k \ge 1$), $G_{IV}$,
  $G_{V}$ as shown in Figure~\ref{fig:forbiddensubgraphs} as subgraphs.
  \label{th:tuckerforbidden}
\end{theorem}

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.35]{../img/tuckersforbiddensubgraphs.png}  
  \caption[\figtabsize Tucker's forbidden subgraphs]{\figtabsize Tucker's
    forbidden subgraphs for convex bipartite graphs. \temptext{ PLACEHOLDER
    IMG }} 
  \label{fig:forbiddensubgraphs}
\end{figure}

Theorem~\ref{th:tuckeratfree} and Theorem~\ref{th:tuckerforbidden}
result in the following Theorem~\ref{th:tuckercop} which characterizes
matrices with \COP.

\begin{theorem}
  (\cite[Th.~9]{at72}, \cite[Th.~2.5]{d08phd})\\
  A matrix $M$ has \COP \iff it contains none of the matrices 
$M_{I_k}$, $M_{II_k}$, $M_{III_k}$ (with $k \ge 1$), $M_{IV}$,
  $M_{V}$ as shown in Figure~\ref{fig:forbiddensubmatrices} as submatrices.
  \label{th:tuckercop}
\end{theorem}

\begin{figure}[htbp]
  \centering
  \includegraphics[scale=0.35]{../img/tuckersforbiddensubmatrices.png}  
  \caption[\figtabsize Tucker's forbidden submatrices]{\figtabsize Tucker's
    forbidden submatrices for convex bipartite graphs. \temptext{ PLACEHOLDER
    IMG }} 
  \label{fig:forbiddensubmatrices}
\end{figure}

It can be verified that the matrices in
Figure~\ref{fig:forbiddensubmatrices} are the half adjacency matrices
of the graphs in Figure~\ref{fig:forbiddensubgraphs} respectively
which is not surprising due to Definition~\ref{def:graphwithcop}~\ref{def::convexbi}.

\temptext{
  \begin{enumerate}[-I-]
  \item PQ trees. Define. Concise flowchart of algorithm? [BL76]\\
    The construction of PQ trees in \cite{bl76} draws on the close
    relationship of matrices with COP to interval graphs. A PQ tree of
    a matrix is one that stores all row (column) permutations of the
    matrix that give the COP orders (there could be multiple orders of
    rows or columns) of the matrix. This is constructed using an
    elaborate linear time procedure and is also a test for
    planarity\tnote{check check check. both interval graph and
      planarity in this paper?}.
  \item Variations - PQR, PC etc. [MM96, Hsu01, McC04]\\
    PQR trees is a generalized data structure based on PQ trees
    \cite{mm96,mpt98}.  \cite{tm05} describes an improved algorithm to
    build PQR
    trees. \tnote{improv in terms of what?}.\\
    Generalized PQ trees McC04 - brief. Will be elaborated in
    Section~\ref{sec:surveycertalgo}\\
    Hsu also invented PC trees \cite{wlh01}\tnote{This result first
      appeared inproc ISAAC92} which is claimed to be much easier to
    implement.
  \item Set theoretic characterizations [Hsu02, NS09]\\
    \cite{wlh02} describes the simpler algorithm for
    COT.\tnote[pressing]{in terms of what?}\\
    \cite{nsnrs09} describes a characterization of consecutive-ones
    property solely based on the cardinality properties of the set
    representations of the columns (rows); every column (row) is
    equivalent to a set that has the row (column) indices of the rows
    (columns) that have one entries in this column (row). This is
    interesting and relevant, especially to this thesis because it
    simplifies COT to a great degree by reducing the solution search
    space owing to the a simple set theoretic characterization.\\
  \item Parallel algorithms [AS95, BS03, CY91]\\
  \item
  \end{enumerate}

}

\tnote[pressing]{ The results in cite~{bl76} on COT are based on the
   result that interval graphs are AT-free chordal graphs.
 -  Tucker's?}%

\tnote[pressing]{cite~{fg65} uses these results to give the first polynomial
  time algorithm for COT. WHICH RESULTS? \ \ \ \ \ check. how do they use it?}%

\tnote[pressing]{ADD: peo exists iff chordal. lexicographic BFS
  [tag:chordalGraph]} %

\clearpage      % Print pending floats
\section{Optimization problems in COP}
\label{sec:surveycopopt}
\tnote{Expand on ref:sec:optcop}

%{\color{blue} \footnotesize
\temptext{
So far we have been concerned about matrices that have the consecutive
ones property. However in real life applications, it is rare that data
sets represented by binary matrices have COP, primarily due to the
noisy nature of data available. At the same time, COP is not arbitrary
and is a desirable property in practical data representation
\cite{co98,jkckv04,k77}. In this context, there are several
interesting problems when a matrix does not have COP but is ``close''
to having COP or is allowed to be altered to have COP. These are the
optimization problems related to a matrix which does not have
COP. Some of the significant problems are surveyed in this section.

@@

\cite{jkckv04} discusses the use of matrices with almost-COP (instead
of one block of consecutive {\un}s, they have $x$ blocks, or {\em
  runs}, of consecutive {\un}s and $x$ is not too large) in the
storage of very large databases.  The problem is that of reordering of
a binary matrix such that the resulting matrix has at most $k$ runs of
{\un}s. This is proved to be NP hard using a reduction from the
Hamiltonian path problem.\tnote{Theorem 2.1 in jkckv} \tnote{(1) A
  connection of COP problem to the travelling salesman problem is also
  introduced. what does this mean? -- COP can be used as a tool to
  reorder $0.5T \le runs(M) \le T$. (2) The optimization version of
  the $k$-run problem, i.e. minimization of number of blocks of ones
  is proven to be NP complete by cite:k77}\tnote{are these two the
  same?} \tnote{what is the reduction?} \tnote{other problems similar
  to COP -- cite:ckl96 (ILP, circ ones, one drop) -- cite:th98
  (generalization of COP - minimax, biotonic column) Tucker}

}

\section{Certifying Algorithms}
\label{sec:surveycertalgo}
%\tnote{Expand on sec:appcertalgo} \tnote{ (certification McC04
%theme)}

\temptext{ Certifying algorithm (Generalized PQ trees) [McC04] -- Move this to section on
    certifying algo.

\cite{mcc04} describes a different approach to COT. While all previous
COT algorithms gave the COP order if the matrix has the property but
exited stating negative if otherwise, this algorithm gives an evidence
by way of a certificate of matrix even when it has no COP. This
enables a user to verify the algorithm's result even when the answer
is negative. This is significant from an implementation perspective
because automated program verification is hard and manual verification
is more viable. Hence having a certificate reinforces an
implementation's credibility. Note that when the matrix {\em has} COP,
the COP order is the certificate.  The internal machinery of this
algorithm is related to the weighted betweenness problem
addressed\tnote{in what way??} in \cite{co98}.  \tnote{expand
  on the COP order graph creation and it having to be bipartite for M
  to have COP. and thus an odd cycle being an evidence of no COP.}

}

\section{COP in Relational Database Model}
\label{sec:surveyrdbm}
\tnote{Expand on sec:apprdbm} \tnote{(set systems theme)}

\section{COP in Graph Isomorphism}
\label{sec:surveygraphiso}
\tnote{Expand on sec:appgraphiso} \tnote{(canonization theme)}

\temptext{
The survey from kklv10 conclusion.
}

% \section{Other applications}
% {\tt If any. \tnote{ where should this go?}: (1) cite|jlm97 (application of PQ
%   trees in graphics). (2) helly's theorem citation
%   19XXdgk-Hellystheorem-Danzer-Gruenbaum-Klee}}

@@

%%%%%%%%%%%%% listing package for pretty formatting BEGIN
% \lstloadlanguages{C}
% \lstset{language=C, commentstyle=\scriptsize,
%   numberstyle=\tiny, numbers=left,               % Line numbers
%   stepnumber=2,numbersep=5pt,firstnumber=10, 
%   showstringspaces=false,                    % No explicit spaces
%   emph={printf},emphstyle=\underbar,      % Specific formatting
%   emph=[2]{sum},emphstyle=[2]\color{blue},      % Sp format (multiple)
%   emph=[3]{i},emphstyle=[3]\color{red}      % Sp format (multiple)
%   }

### 2012-01-05 src/040chap4.tex
Chapter 2 edited. See diff for details. Too many to remember. or maybe too little.

>>>>

\tnote[pressing]{TBD survey
  see blue notes (in notebook) under Graph Isomorphism. namely
  citations in cite:aas93 (3) a brief on heirarchy of chordal
  graphs, path graphs, interval graphs, peo, clique tree etc. and
  results. (4) interval graphs are incomparability graphs. see
  Golumbic. [tag:classification] (5) finding min length hole in
  bipartite graph is polynomial time [Sec 3.3.2 in cite:d08phd] (6)
  theorem 2.2 in cite:d08phd - $G$ is union of v.d. caterpillars iff
  $M$ has COP in rows. $M$ is edge vertex incidence matrix $(*,2)$
  of $G$
}%
